Lossy compression of DNA sequencing quality data

### 2015-05-13 manuscript/q-compression.tex
Added reference for Burrows-Wheeler algorithm, which is the algorithm in BZip2. Revised the description of economic costs for all methods, and added cost description for QualComp. Described the advantageous properties of profile-based compression with regard to increases in either or both read length and number of reads. Changed figure legend to reflect the use of arrows. Corrected a spelling error, and made some other small revisions.

The economic costs of memory use for binning, in general terms,
include no fixed costs, and marginal costs that are a function of the
number of base-call quality values times the cost of the encoding.

@@

Although we focus only on two bins in this work, a greater number of
bins can be used in practice, albeit with some higher cost in memory.
For example, an additional third bin, thus giving bins of good, bad,
and intermediate, could be used for base qualities near the border
between bins in the 2-bin encoding. Examining the trade-off of
additional bins and their resulting compressibility and effect on
downstream analyses are not considered here.

@@

The economic costs of memory use for model-based compression, in
general terms, include no fixed costs, and marginal costs that are a
function of the number of reads times the cost of representing the
model parameters.

@@

The economic costs of memory use for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is a function of the number of profiles times the cost
of encoding them, and these fixed costs are amortized over the entire
set of reads to which they apply. Additionally there are marginal
costs that are a function of the number of reads encoded times the
cost of the encoding.  Profiling like modeling, but not binning, has
the advantageous property that cost improves with read length.

Profiles can be obtained from techniques other than $k$-means
clustering. For example, we could store the profiles of polynomial
functions to increase the compressibility of polynomial regression
modeling.  In other words, we could use a spline (a function that is
piecewise-defined by polynomial functions) to represent a given
quality sequence.  However, we have not explored these other
approaches here.

@@

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool~\citep{Ochoa:2013rt}. The program models
quality values as a multivariate Gaussian distribution, computing the
mean and covariance for each position in the read, and these varlues
are stored by the decoder to later reconstruct a \emph{representative}
quality value. \textsc{QualComp} takes as input a user-specified rate
(bits/read), and these bits are apportioned among positions within the
read to minimize the average error. The quality values can be
clustered beforehand to produce more accurate models.

@@

As a measure of compression effectiveness we use bits/base-call, and
define it as the size of the compressed representation of quality
values (in bits) divided by the number of quality values represented.
As a measure of information loss we use mean squared error
(\textsc{mse}) as a loss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
sequences, $Q_i'$ is the compressed/decompressed quality value, and
$Q_i$ is the original quality value associated with sequence position
$i$.

@@

\begin{figure*}[!tb]
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods and subsequently further
  compressed using a general algorithm via \textsc{bz}ip2 to four data
  sets: \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of
  $n$. Asterisks denote the corresponding lossless compression using
  \textsc{bz}ip2, with the black asterisk corresponding to original
  data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
different compressed representations, which include both fixed and
marginal costs, and subsequently further compressed using a general
algorithm via \textsc{bz}ip2 (Figure \ref{fig:mse_vs_bpbp}). Here we
present the fragment libraries for the \textit{Rhodobacter
  sphaeroides}, and \textit{Homo sapiens} data sets; the short-jump
library results are available in Supplemental Table 1. Storing the
uncompressed quality values requires 1 byte/base-call because they are
stored in \textsc{ascii} format, and the lossless compression of each
data set using \textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call;
these values provide a reference point using a widely available
compression program. Values from the same class of compression method
tend to cluster together across the different data sets, with the
0-degree polynomial regression, profile encodings, and
\textsc{q}ual\textsc{c}omp have the lowest bits/base-call.

>>>>

The economic costs of memory use for binning, in general terms,
include no fixed costs, and marginal costs that are the product of the
number of base-call quality values times the cost of the encoding.

@@

Although we focus only on two bins in this work, a greater number of
bins can be used in practice, albeit with a higher cost in memory
requirement.  For example, an additional third bin, thus giving bins
of good, bad, and intermediate, could be used for base qualities near
the border between bins in the 2-bin encoding. Additional bins beyond
two, and their resulting compressibility and effect on downstream
analyses, are not considered here.

@@

The economic costs of memory use for model-based compression, in
general terms, include no fixed costs, and marginal costs that are the
product of the number of reads times the cost of representing the
model parameters.

@@

The economic costs of memory use for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is the product of the number of profiles times the
cost of encoding them, and these fixed costs are amortized over the
entire set of reads to which they apply. Additionally there are
marginal costs that are the product of the number of reads encoded
times the cost of the encoding. Profiling like modeling, but not
binning, has the advantageous property that marginal costs are the
reciprocal of the read length. Thus the costs of memory use for
profile-based compression decrease proportionally with increases in
either or both the number of reads and read length.

Profiles can be obtained from techniques other than $k$-means
clustering. For example, we could store the profiles of polynomial
functions to increase the compressibility of polynomial regression
modeling. In other words, we could use a spline (a function that is
piecewise-defined by polynomial functions) to represent a given
quality sequence. However, we have not explored this or other
approaches here.

@@

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool~\citep{Ochoa:2013rt}. The program models
quality values as a multivariate Gaussian distribution, computing the
mean and covariance for each position in the read, and these values
are stored by the decoder to later reconstruct a \emph{representative}
quality value. \textsc{QualComp} takes as input a user-specified rate
(bits/read), and these bits are apportioned among positions within the
read to minimize the average error. The quality values can be
clustered beforehand to produce more accurate models.

The economic costs of memory use for compression with
\textsc{QualComp}, in in our interpretation, appear to have no fixed
costs, and the marginal costs that are the product of the number of
reads times the cost of representing the model parameters.

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs after subsequent compression using the
Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, we use
bits/base-call, define it as the size of the compressed representation
of quality values (in bits) divided by the number of quality values
represented. As a measure of information loss we use mean squared
error (\textsc{mse}) as a loss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
sequences, $Q_i'$ is the compressed/decompressed quality value, and
$Q_i$ is the original quality value associated with sequence position
$i$.

@@

\begin{figure*}[!tb]
\centerline{\includegraphics[width=7in]{compression_results.pdf}}
\caption{The relationship of bits/base-call and mean squared error for
  quality value compression methods applied to four data sets:
  \textit{Rhodobacter sphaeroides} 2.4.1; \textit{Homo sapiens}
  chromosome 14; \textit{Escherichia coli} str. K-12 MG1655; and
  \textit{Mus musculus}. Point labels correspond to different
  compression methods: 2B, 2-bin encoding; P$n$, profiling with $n$
  profiles; R$n$, modeling with polynomial regression of degree $n$;
  Q$n$, \textsc{q}ual\textsc{c}omp with rate parameter of $n$. Arrows
  denote the corresponding lossless compression using \textsc{bz}ip2,
  with the black arrow corresponding to original data.}
\label{fig:mse_vs_bpbp}
\end{figure*}

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
different compressed representations (Figure
\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets;
the short-jump library results are available in Supplemental Table
1. Storing the uncompressed quality values requires 1 byte/base-call
because they are stored in \textsc{ascii} format, and the lossless
compression of each data set using \textsc{bz}ip2 ranges from
2.19--3.10 bits/base-call; these values provide a reference point
using a widely available compression program. Values from the same
class of compression method tend to cluster together across the
different data sets, with the 0-degree polynomial regression, profile
encodings, and \textsc{q}ual\textsc{c}omp have the lowest
bits/base-call.
